En garde! Unguarded iteration for reversible computation in the Delay monad\thanks{Niccol\`o Veltri was supported by a research grant (13156) from VILLUM FONDEN.}

### 2019-04-30 mpc19/paper.tex
conclusion on other monads of partiality

% Delay macros
\newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
\newcommand{\Maybe}{\mathsf{Maybe}}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{unfolding}}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}
\newcommand{\daggerD}{\mathsf{dagger}_\D}
\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
\newcommand{\done}{\mathsf{done}}
\renewcommand{\next}{\mathsf{next}}
\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}

@@

Weak bisimilarity is defined in terms of convergence. A computation
$x : \Delay A$ converges to $a : A$ if it terminates in a finite
number of steps returning the value $a$. When this happens, we write
$x \dn a$. The relation $\dn$ is inductively defined by the rules:
\[
\infer{\now\,a \dn a}{}
\quad
\infer{\later\,x \dn a}{x \dn a}
\]
Two computations in $\Delay A$ are weakly bisimilar if they differ by
a finite number of applications of the constructor
$\later$. Alternatively, we can say that two computations $x$ and $y$
are weakly bisimilar if, whenever $x$ terminates returning a value
$a$, then $y$ also terminates returning $a$, and vice versa. This
informal statement can be formalized in several different but
logically equivalent ways \cite{ChapmanUV19}. Here we consider a
coinductive formulation employed e.g. in \cite{Danielsson18}.
%% A possible defintion is as follows:
%% \begin{equation}\label{eq:weak1}
%% x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
%% \end{equation}
%% The definition says that $x$ and $y$ are weakly bisimilar if, whenever
%% $x$ terminates returning a value $a$, then $y$ also terminates
%% returning $a$, and vice versa.
%% Alternatively, weak bisimilarity can be coinductively formulated as follows:
\begin{equation}
\label{eq:weak2}
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{cc}
\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
&
\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
\\
\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
&
\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
\end{array}
\end{equation}
Notice that the constructor $\later_{\approx}$ is coinductive. This
allows us to prove $\never \approx \never$. 
%% Formulation
%% (\ref{eq:weak1}) and (\ref{eq:weak2}) are logically equivalent,
%% meaning that two computations are weakly bisimilar as in
%% (\ref{eq:weak1}) if and only if they are weakly bisimilar as in
%% (\ref{eq:weak2}).  In the rest of the paper we will employ both
%% encodings of weak bisimilarity, in each case explicitly stating if we
%% are working with definition (\ref{eq:weak1}) or definition
%% (\ref{eq:weak2}). 
 Weak bisimilarity is an equivalence relation and it
is a congruence w.r.t. the $\later$ operation.  For example, here is a
proof that weak bisimilarity is reflexive.
\begin{align*}
& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
\end{align*}

@@

As an alternative to quotienting the homsets of $\D$, we could have
quotiented the delay datatype by weak bisimilarity:
$\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$. In previous work
\cite{ChapmanUV19}, we showed that this construction has problematic
consequences if we employ Hofmann's approach to quotient types
\cite{Hofmann}. For example, it does not seem possible to lift the
monad structure of $\Delay$ to $\mathsf{Delay}_{\approx}$ without
postulating additional principles such as the axiom of countable
choice. More fundamentally for this work, countable choice would be
needed for modelling the trace operator of \Pio\ in the Kleisli
category of $\mathsf{Delay}_{\approx}$. Notice that, if the setoid approach
to quotienting is employed, the latter constructions go through
without the need for additional assumptions. In order to keep an
agnostic perspective on quotient types and avoid the need for
disputable semi-classical choice principles, we decided to quotient
the homsets of $\D$ by (poitwise) weak bisimilarity instead of the
delay datatype.

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. The
binary product of $A$ and $B$ in $\Dapprox$ is given by
$A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
$- \times -$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying a number of axioms, most notably the unfolding axiom:
\[
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the unfolding
axiom. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

>>>>

% Delay macros
\newcommand{\Delay}{\ensuremath{\mathsf{Delay}\,}}
\newcommand{\Maybe}{\mathsf{Maybe}}
\newcommand{\now}{\mathsf{now}}
\newcommand{\later}{\mathsf{later}}
\newcommand{\laterR}{\mathsf{laterR}}
\newcommand{\laterL}{\mathsf{laterL}}
\newcommand{\never}{\mathsf{never}}
\newcommand{\dn}{\downarrow}
\newcommand{\bind}{\mathsf{bind}}
\newcommand{\str}{\mathsf{str}}
\newcommand{\costr}{\mathsf{costr}}
\newcommand{\D}{\mathbb{D}}
\newcommand{\Dapprox}{\mathbb{D}_{\approx}}
\newcommand{\copairD}[2]{[#1,#2]_{\D}}
\newcommand{\pairD}[2]{\left< #1,#2 \right>_{\D}}
\newcommand{\inlD}{\mathsf{inl}_{\D}}
\newcommand{\inrD}{\mathsf{inr}_{\D}}
\newcommand{\fstD}{\mathsf{fst}_{\D}}
\newcommand{\sndD}{\mathsf{snd}_{\D}}
\newcommand{\piso}{\mathsf{isPartialIso}}
\newcommand{\pisoalt}{\mathsf{isPartialIso}_2}
\newcommand{\iterD}{\mathsf{iter}_\D}
\newcommand{\unfolding}{\mathsf{fixpoint}_\D}
\newcommand{\refl}{\mathsf{refl}}
\newcommand{\traceD}{\mathsf{trace}_\D}
\newcommand{\traceHD}{\mathsf{traceH}_\D}
\newcommand{\daggerD}{\mathsf{dagger}_\D}
\newcommand{\Orb}[3]{\mathsf{Orb}\,#1\,#2\,#3}
\newcommand{\done}{\mathsf{done}}
\renewcommand{\next}{\mathsf{next}}
\newcommand{\reverseOrbit}{\mathsf{reverseOrb}}

@@

Weak bisimilarity is defined in terms of convergence. A computation
$x : \Delay A$ converges to $a : A$ if it terminates in a finite
number of steps returning the value $a$. When this happens, we write
$x \dn a$. The relation $\dn$ is inductively defined by the rules:
\[
\infer{\now\,a \dn a}{}
\quad
\infer{\later\,x \dn a}{x \dn a}
\]
Two computations in $\Delay A$ are weakly bisimilar if they differ by
a finite number of applications of the constructor
$\later$. Alternatively, we can say that two computations $x$ and $y$
are weakly bisimilar if, whenever $x$ terminates returning a value
$a$, then $y$ also terminates returning $a$, and vice versa. This
informal statement can be formalized in several different but
logically equivalent ways \cite{ChapmanUV19,VeltriPhD}. Here we
consider a coinductive formulation employed e.g. in
\cite{Danielsson18}.
%% A possible defintion is as follows:
%% \begin{equation}\label{eq:weak1}
%% x \approx y = (a : A) \to x \dn a \leftrightarrow y \dn a
%% \end{equation}
%% The definition says that $x$ and $y$ are weakly bisimilar if, whenever
%% $x$ terminates returning a value $a$, then $y$ also terminates
%% returning $a$, and vice versa.
%% Alternatively, weak bisimilarity can be coinductively formulated as follows:
\begin{equation}
\label{eq:weak2}
\arraycolsep=8pt\def\arraystretch{2.8}
\begin{array}{cc}
\infer{\now_{\approx}  : \now\,a \approx \now\,a}{}
&
\infer={\later_{\approx} \,p : \later\,x_1 \approx \later\,x_2}{p : x_1 \approx x_2}
\\
\infer{\laterL_{\approx} \, p: \later\,x \approx \now\,a}{p : x \approx \now \,a}
&
\infer{\laterR_{\approx} \, p: \now\,a \approx \later\,x}{p : \now\,a \approx x}
\end{array}
\end{equation}
Notice that the constructor $\later_{\approx}$ is coinductive. This
allows us to prove $\never \approx \never$. 
%% Formulation
%% (\ref{eq:weak1}) and (\ref{eq:weak2}) are logically equivalent,
%% meaning that two computations are weakly bisimilar as in
%% (\ref{eq:weak1}) if and only if they are weakly bisimilar as in
%% (\ref{eq:weak2}).  In the rest of the paper we will employ both
%% encodings of weak bisimilarity, in each case explicitly stating if we
%% are working with definition (\ref{eq:weak1}) or definition
%% (\ref{eq:weak2}). 
 Weak bisimilarity is an equivalence relation and it
is a congruence w.r.t. the $\later$ operation.  For example, here is a
proof that weak bisimilarity is reflexive.
\begin{align*}
& \refl_{\approx} : \{x : \Delay A\} \to x \approx x \\
& \refl_{\approx} \,\{\now\,a\} = \now_\approx \\
& \refl_{\approx} \,\{\later\,x\} = \later_\approx\,(\refl_\approx \,\{x\})
\end{align*}

@@

As an alternative to quotienting the homsets of $\D$, we could have
quotiented the delay datatype by weak bisimilarity:
$\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$. In previous work
\cite{ChapmanUV19}, we showed that this construction has problematic
consequences if we employ Hofmann's approach to quotient types
\cite{Hofmann}. For example, it does not seem possible to lift the
monad structure of $\Delay$ to $\mathsf{Delay}_{\approx}$ without
postulating additional principles such as the axiom of countable
choice. More fundamentally for this work, countable choice would be
needed for modelling the trace operator of \Pio\ in the Kleisli
category of $\mathsf{Delay}_{\approx}$. Notice that, if the setoid approach
to quotienting is employed, the latter constructions go through
without the need for additional assumptions. In order to keep an
agnostic perspective on quotient types and avoid the need for
disputable semi-classical choice principles, we decided to quotient
the homsets of $\D$ by (pointwise) weak bisimilarity instead of the
delay datatype.

@@

Just as limits in $\Set$ do not lift to limits in the category $\Par$ of sets
and partial functions, they do not lift to $\Dapprox$ either. This is not an
issue with these concrete formulations of partiality, but rather with the
interaction of partiality (in the sense of restriction categories) and limits
in general (see \cite[Section 4.4]{CockettL07}). In particular, $1$ is not
terminal object and $A \times B$ is not the binary product of $A$ and $B$ in
$\Dapprox$. In fact, $0$ is (also) the terminal object, with $\lambda \_.\,
\never : A \to \Delay 0$ as the terminal morphism. The
binary product of $A$ and $B$ in $\Dapprox$ is given by
$A + B + A \times B$. Nevertheless, it is possible to prove that 1 and
$\times$ are partial terminal object and partial binary products
respectively, in the sense of Cockett and Lack's restriction
categories \cite{CockettL02,CockettL07}. Here we refrain from making the latter
statement formal. We only show the construction of the partial pairing
operation, which we employ in the interpretation of \Pio.  Given
$f : C \to \Delay A$ and $g : C \to \Delay B$, we define:
\begin{align*}
& \pairD f g : C \to \Delay (A \times B) \\
& \pairD f g = \costr \diamond (\str \comp \pair f g)
\end{align*}
Since the delay monad is commutative, the function $\pairD f g$ is
equal to $\str \diamond (\costr \comp \pair f g)$. The operation
$\pairD - -$ is compatible with weak bisimilarity.

@@

An Elgot monad \cite{GoncharovMR16,GoncharovSRJ18} is a monad $T$
whose Kleilsi category supports unguarded uniform iteration. More
precisely\footnote{Here we give the definition of Elgot monad on
  $\Set$, but the definition of Elgot monad makes sense in any
  category with binary coproducts.}, a monad $\T$ is Elgot if there
exists an operation
\[
\iter_\T : (A \to \T \,(B + A)) \to A \to \T\,B
\]
satisfying a number of axioms:
\[\arraycolsep=8pt\def\arraystretch{1.8}
\begin{array}{c}
\mathsf{fixpoint} :
\iter_\T\,f \equiv \copair {\eta_\T}{\iter_\T \,f} \diamond_\T f
\\
\mathsf{naturality} : g \diamond_\T \iter_\T\,f \equiv
  \iter_\T\,(\copair{\T \inl \comp g} {\eta \comp \inr} \diamond_\T f)
\\
\mathsf{codiagonal} :
\iter_\T \,(\iter_\T\,g) \equiv \iter_\T\,(\T \copair {\id} {\inr}
  \comp g)
\\[.25cm]
\infer{\mathsf{uniformity}\,p : \iter_\T \,f \comp h \equiv \iter_\T
  \,g}{p : f \comp h \equiv \T (\id + h) \comp g}
\end{array}
\]
where $\eta_\T$ is the unit of $\T$ and $\diamond_\T$ denotes morphism
composition in the Kleisli category of $\T$.
The standard definition of uniform iteration operator includes the
dinaturality axiom, which has recently been discovered to be derivable
from the other laws \cite{EsikG16,GoncharovMR16}.

@@

As mentioned above, $\iterD$ satisfies the Elgot iteration axioms only
up to weak bisimilarity. Here we show the proof of the fixpoint
axiom. The proof relies on an auxiliary proof $\unfolding'$. 
\[
\def\arraystretch{1.2}
\begin{array}{ll}
\multicolumn{2}{l}{\unfolding' : (f : A \to \Delay (B + A)) \to
  \bind\, \copairD {\now}{\iterD \,f} \approx \iterD' \,f} \\
\unfolding' \,f \,(\now \,(\inl \,b)) & = \now_\approx \\
\unfolding' \,f \,(\now \,(\inr \,a)) & = \laterR_{\approx} \,
  \refl_\approx \\
\unfolding' \,f \,(\later \,x) & = \later_\approx \,(\unfolding'\, f\,
  x) \\
 \\
\multicolumn{2}{l}{\unfolding : (f : A \to \Delay (B + A)) \to
  \copairD {\now}{\iterD \,f}\diamond f \approx \iterD \,f}\\
\multicolumn{2}{l}{\unfolding \,f\,x = \unfolding'\,f\,(f\,x)}
\end{array}
\]

@@

It is natural to wonder if our work can be ported to other monads of
partiality in Martin-L\"of type theory. As already discussed in the
end of Section \ref{sec:trace}, the maybe monad is not suitable for
modelling a well-behaved trace combinator without the assumption of
classical principles such as \textsf{LPO}. The partial map classifier
\cite{UustaluV17-2,EscardoK17}
$\mathsf{PMC}\,A = (P : \mathsf{Prop}) \to P \to A$, where
$\mathsf{Prop}$ is the type of propositions (types with at most one
inhabitant), supports the existence of a uniform iteration operator
and therefore a trace. Nevertheless, the specification of iteration is
more complicated then the one presented in Section \ref{sec:elgot} for
the delay monad, which is a simple corecursive definition. The Elgot
monad structure of $\mathsf{PMC}$ follows from its Kleisli category
being a join restriction category, so iteration is defined in terms of
least upper bounds of certain chains of morphisms. The subcategory of
partial isomorphisms of the Kleisli category of $\mathsf{PMC}$
supports a dagger trace combinator, which can be proved following
Kaarsgaard et al.'s general strategy \cite{KaarsgaardAG17}.  The exact
same observations apply to the partiality monad in homotopy type
theory \cite{AltenkirchDK17,ChapmanUV19}, to which the quotiented
delay monad $\mathsf{Delay}_{\approx}\,A = \Delay A/{\approx}$ is
isomorphic to under the assumption of countable choice.
